---
title: "Introduction to typesys"
author: "Nick Ulle"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to typesys}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo = FALSE}
library(typesys)
```

The __typesys__ package formalizes R's type system into a simple language.

The most basic expressions in the type language are constant types, such as
`Integer` and `Logical`. Unlike R's `typeof()` function, we distinguish between
scalars and vectors. Constant types always describe scalar values. For example,
`Integer` is a scalar integer. Generally, the names of the constant types match
the names returned by `typeof()` up to capitalization. There are two
exceptions:

1. An R "double" corresponds to the `Numeric` type.
1. An R "character" corresponds to the `String` type.

The names `Numeric` and `String` were chosen because they give clearer
descriptions of the values that are actually in their respective types.

If we want to create a new `Integer` type, we can write
```{r}
IntegerType()
```
The output of this function is an S4 class that represents an `Integer`.
Similar constructor functions exist for every other constant type.

The type language also includes type variables, which are named placeholders
for unknown types. The constructor for type variables requires a name to be
specififed. To create a type variable called `a`, we can write
```{r}
TypeVariable("a")
```
Type variables serve two purposes. First, some functions have parameters where
more than one type is allowed. For instance, the identity function
```{r, results = "hide"}
function(x) x
```
accepts an argument of any type. These functions are called polymorphic
functions. Type variables are necessary to express the type of a polymorphic
function. This is discussed in more detail below. Second, type variables are
useful in any application where types may not be unknown. Type inference is an
example application.

<!--
Vectors have certain properties, such as length, regardless of what the type of
the individual elements is. Because of this, vectors are a parameterized type
`Vector(a)` in the type language. The parameter `a` is the element type. If we
want to create a new integer vector type, we can write
```{r}
#VectorType(IntegerType())
```
Similarly, to create a new character vector type, we can write
```{r}
#VectorType(StringType())
```

Functions are another parameterized type. They have the form $a \to b$ for
arguments $a$ (a list of types) and return type $b$.
-->


## The Formula-based Type Language

Constructing expressions in the type language using the type constructor
functions is more verbose than is really necessary. __typesys__ provides a way
to write type expressions concisely using R's formula notation. The rules are:

*   A formula with both sides is converted to a function.
*   A formula with right-hand side only is converted to the type described on
    the right-hand side.
*   The suffix "Type" is omitted from type constructors.
*   Constant types are constructed without parentheses.
*   Type variables are implicitly constructed from lowercase names (constant
    types always use title case).
*   Tuple and record types are constructed with `c()`, similar to R's syntax.

Thus we can express the type $(a, b) \to Integer$ in formula notation as
```{r}
exp = c(a, b) ~ Integer
formula_to_type(exp)
```
Here the function `formula_to_type()` converts the formula to a type. The type
variables can optionally be quantified by setting the function's `quantify`
parameter.


## Unification & Substitution

__typesys__ also includes an implementation Robinson's unification algorithm in
the `unify()` function. Given two algebraic expressions, unification attempts
to find a substitution of variables that makes the two expressions equal.

For example, suppose we have types $a \to Integer$ and $Numeric \to Integer$
and expect that they are equal. Then we can unify these types to see that they
are equal under the substitution $a \mapsto Numeric$. We would
write this as
```{r}
x = formula_to_type(a ~ Integer)
y = formula_to_type(Numeric ~ Integer)
unify(x, y)
```

The function emits an error when two types cannot be unified because they are
incompatible.
```{r, error = TRUE}
x = formula_to_type(a ~ Integer)
y = formula_to_type(~ Numeric)
unify(x, y)
```

Successful unification returns a substitution of variables. Substitutions can
be applied to existing types to with the `do_substitution()` function. For
instance, by unifying the type variable `a` and an integer gives the
substitution $a \mapsto Integer$. We can then apply the substitution to any
expression that contains `a` to replace `a` with `Integer`. The code for this
is
```{r}
x = TypeVariable("a")
y = IntegerType()
sub = unify(x, y)

do_substitution(x, sub)
```

Finally, note that the `unify()` function is an inefficient prototype. In the
worst case, it requires exponential time in the number of subexpressions being
unified. Better algorithms for unification are known, with the best running in
linear time. In the future, the `unify()` function may be changed to use a
better algorithm.
